---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RainAttr

<!-- badges: start -->
<!-- badges: end -->

`RainAttr` provide tools for estimation and inference of attribution and sample average treatment effect in rainfall enhancement trials, based on the two-stage linear mixed model approach employed by [Chambers et al. (2022)](#references). It also provides tools for exploratory analysis of rainfall enhancement trial data.

## Installation

You can install the development version of RainAttr from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("Zy1225/RainAttr")
```

## Example

This basic example shows how to use the package for estimation, bootstrap inference, and permutation inference.

```{r example}
library(RainAttr)

result = rain_attr(
  #gauge-day level trial data
  data = oman,
  
  #LMM formula fitted to the upwind (first-stage) observations
  upwind_lmm_formula = LogRain ~  Gauge.Elevation + Steering.Wind.Speed + Total.Totals + PC2.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure + (1|TrialDay),
  
  #Variable name for storing instrumental prediction from upwind (first-stage) LMM
  instr_pred_name = 'natural_pred',
  
  #Types of instrumental prediction: 'Unconditional' or 'Conditional'
  instr_pred_type = 'Conditional',
  
  #LMM formula fitted to the downwind (second-stage) observations
  downwind_lmm_formula = LogRain - natural_pred ~ Gauge.Elevation  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02 + (1|TrialDay),
  
  #Formula for fitting a logistic regression model to the indicator of rainfall event (used for bootstrapping of rainfall events when bootstrap_opt(bootstrap_zero) = TRUE)
  downwind_logistic_formula = (Rain.Gauge.Measurement > 0) ~ Gauge.Elevation + natural_pred  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02,
  
  #Formula for fitting the propensity score model to downwind observations, where responses are indicators of whether each downwind observations is a 'Target'
  downwind_propensity_formula = (Gauge.Day.Type == 'Target') ~ Total.Totals + PC1.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure,
  
  #Specify the column in input data that contains the raw rainfall
  rain_col_name = 'Rain.Gauge.Measurement',
  
  #Logical expression identifying subset of observations to which upwind_lmm_formula is fitted
  upwind_subset = Gauge.Day.Type == 'Upwind',

  #Logical expression identifying subset of observations to which downwind_lmm_formula is fitted
  downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),

  #Logical expression identifying subset of observations of those that are 'Target'
  downwind_target_subset = Gauge.Day.Type == 'Target',

  #Logical expression identifying subset of observations of those that are 'Control'
  downwind_control_subset = Gauge.Day.Type == 'Control',

  #Logical expression identifying subset of observations with rainfall event
  positive_subset = Rain.Gauge.Measurement > 0,
  
  #Types of correction (for back-transformation bias) used for computing the attribution estimate: 'ChambersEtAl', 'ChambersEtAl_No_Winsorize', 'ThoEtAl', or 'No'
  attr_type = 'ThoEtAl',

  #Vector of variable names in downwind_lmm_formula to identify non-ionizer related covariates, whose effects are not included in the calculation of attribution and SATE
  x_downwind_name = c('Gauge.Elevation'),
  
  #Indicator for whether to perform bootstrap inference on attribution and SATE
  bootstrap = TRUE,

  #Specification of various option for bootstrap, e.g., bootstrap_opt(bootstrap_type = 'PREB1'), currently support bootstrap_type = 'PREB0', 'PREB1', 'PREB2', 'REB0', 'REB1', 'REB2', 'MREB1'
  bootstrap_option = bootstrap_opt(B_bootstrap = 500, 
                                   bootstrap_seed = 123, 
                                   bootstrap_parallel = TRUE,
                                   bootstrap_parallel_num_worker = 6),

  #Indicator for whether to perform permutation inference on attribution and SATE
  permutation = TRUE,

  #Specification of various option for permutation,  e.g., whether to permute the operating states between ionizers, between days, between gaugedays
  permutation_option = permutation_opt(B_permutation = 500,
                                       permutation_seed = 999,
                                       permutation_parallel = TRUE,
                                       permutation_parallel_num_worker = 6)
  )
```

The main function, `rain_attr`, returns an S3 object that supports a range of common S3 methods, including:

```{r summary}
summary(result)
```

and

```{r out.width="100%", fig.height= 12}
plot(result)
```

## Documentation

Full documentation, function references, and worked examples are available on
the [package website](https://zy1225.github.io/RainAttr/).

## Citation

To cite `RainAttr`, run:

```{r}
citation("RainAttr")
```

## References

Chambers, R., Ranjbar, S., Salvati, N., and Pacini, B. (2022). Weighting, informativeness and causal inference, with an application to rainfall enhancement. *Journal of the Royal Statistical Society: Series A (Statistics in Society)*, **185**, 1584–1612.
