---
title: "Introduction to RainAttr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RainAttr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(RainAttr)
devtools::load_all(".")
```



## Using the package
This package estimates and performs inference on the **attribution**, defined as the percentage increase in raw-scale rainfall attributable to the ionization technology, and **sample average treatment effect (SATE)** for gauge-day level rainfall enhancement trial data. An example of a gauge-day level dataset is given below:

```{r}
head(oman)
```

The `Gauge.Day.Type` column classifies each observation into one of the following types:

* Upwind: gauges located upwind of ionizers
* Target: gauges located downwind of at least one active ionizers
* Control: gauges located downwind of inactive ionizers
* Out of Scope: gauges that are neither upwind, target, nor control


### Example 

#### Using the main function  

We examine the effectiveness of the ionization technology used in the Oman 2013-2018 trial, based on the two-stage linear mixed model (LMM) approach employed in Chambers et al. (2022a). This is achieved by using the main function `rain_attr` from the package:

```{r}
result = rain_attr(
  #gauge-day level trial data
  data = oman,

  #LMM formula fitted to the upwind (first-stage) observations
  upwind_lmm_formula = LogRain ~  Gauge.Elevation + Steering.Wind.Speed + Total.Totals + PC2.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure + (1|TrialDay),

  #Variable name for storing instrumental prediction from upwind (first-stage) LMM
  instr_pred_name = 'natural_pred',

  #Types of instrumental prediction: 'Unconditional' or 'Conditional'
  instr_pred_type = 'Conditional',

  #LMM formula fitted to the downwind (second-stage) observations
  downwind_lmm_formula = LogRain - natural_pred ~ Gauge.Elevation  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02 + (1|TrialDay),

  #Formula for fitting a logistic regression model to the indicator of rainfall event (used for bootstrapping of rainfall events)
  downwind_logistic_formula = NULL,

  #Formula for fitting the propensity score model to downwind observations, where responses are indicators of whether each downwind observations is a 'Target'
  downwind_propensity_formula = (Gauge.Day.Type == 'Target') ~ Total.Totals + PC1.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure,

  #Specify the column in input data that contains the raw rainfall
  rain_col_name = 'Rain.Gauge.Measurement',

  #Logical expression identifying subset of observations to which upwind_lmm_formula is fitted
  upwind_subset = Gauge.Day.Type == 'Upwind',

  #Logical expression identifying subset of observations to which downwind_lmm_formula is fitted
  downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),

  #Logical expression identifying subset of observations of those that are 'Target'
  downwind_target_subset = Gauge.Day.Type == 'Target',

  #Logical expression identifying subset of observations of those that are 'Control'
  downwind_control_subset = Gauge.Day.Type == 'Control',

  #Logical expression identifying subset of observations with rainfall event
  positive_subset = Rain.Gauge.Measurement > 0,
  
  #Types of correction (for back-transformation bias) used for computing the attribution estimate: 'ChambersEtAl', 'ChambersEtAl_No_Winsorize', 'ThoEtAl', or 'No'
  attr_type = 'ThoEtAl',

  #Vector of variable names in downwind_lmm_formula to identify non-ionizer related covariates, whose effects are not included in the calculation of attribution and SATE
  x_downwind_name = c('Gauge.Elevation')
  )
```

The `rain_attr` function fits two LMMs sequentially:

1. An upwind (first-stage) LMM fitted using `lme4::lmer(upwind_lmm_formula)` to `data[upwind_subset & positive_subset,]` .
    a) Obtain fitted values (named as `instr_pred_name`) for `data[downwind_subset & positive_subset,]`, where the fitted values are conditional or unconditional depending on `instr_pred_type'.
2. A downwind (second-stage) LMM fitted using `lme4::lmer(downwind_lmm_formula)` to `data[downwind_subset & positive_subset,]`.

The estimated LMM model parameters are then used to calculate attribution and SATE estimates, where `attr_type` specifies the type of correction for back-transformation bias in computing the attribution estimates, and `x_downwind_name` identifies non-ionizer related covariates, whose effects are excluded from the calculation of these estimates.

Moreover, a logistic propensity score model is also fitted using `glm(downwind_propensity_formula, family = "binomial")` to `data[downwind_subset & positive_subset,]` with the response being the binary indicator for whether each observation is exposed to the ionizer (i.e., a target observation). This fitted model, along with the arguments specified via `downwind_target_subset` and `downwind_control_subset`, facilitate the estimation of SATEs.



#### Returned object from the main function

The `rain_attr` function returns an S3 object of class `rain_attr`, which can be called to print the main results associated with the attribution and SATEs. Some basic information related to the estimated upwind (first-stage) and downwind (second-stage) models are also printed, such as the number of observations and the estimated fixed effects.
```{r}
result
```

The `apo` and `apl` attribution estimates indicate that `r round(result$hatattr$apo * 100,2)`% and `r round(result$hatattr$apl * 100,2)`% increase in raw-scale downwind rainfall are attributable to the operation of the ionizers, expressed as a percentage of the total observed rainfall and the total natural rainfall, respectively. We note that attribution estimates are only meaningful when log-rainfall is modeled in both stages, which is typically the case since raw-scale rainfall is often heavily right-skewed, whereas log-rainfall better satisfies the normality assumptions of LMMs.

The above results also show that all SATE estimates are positive, suggesting a causal effect of ionization technology in increasing downwind rainfall. For exact formulas of the five types of SATE estimates, see `help(rain_attr)`. A detailed comparison is also provided in Chambers et al. (2022b).



#### Bootstrap and Permutation Inference
The `rain_attr` function also allows for bootstrap and permutation inference on the attribution and SATEs, by setting the arguments `bootstrap = TRUE` and `permutation = TRUE`. Additional options for bootstrap and permutation can be set by inputting named lists to the arguments `bootstrap_option` and `permutation_option`, with the default options returned by `bootstrap_opt()` and `permutation_opt()`. For example, `bootstrap_option = bootstrap_opt(B_bootstrap = 500)` and `permutation_option = permutation_opt(B_permutation = 500)` set the number of bootstrap replicates and permutation replicates to be 500.

Here is an example of performing bootstrap and permutation inference, each with 500 replicates:
```{r, eval=FALSE}
boot_perm_result = rain_attr(
  data = oman,
  upwind_lmm_formula = LogRain ~  Gauge.Elevation + Steering.Wind.Speed + Total.Totals + PC2.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure + (1|TrialDay),
  instr_pred_name = 'natural_pred',
  instr_pred_type = 'Conditional',
  downwind_lmm_formula = LogRain - natural_pred ~ Gauge.Elevation  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02 + (1|TrialDay),
  
  #Formula for fitting a logistic regression model to the indicator of rainfall event (used for bootstrapping of rainfall events when bootstrap_opt(bootstrap_zero) = TRUE)
  downwind_logistic_formula = (Rain.Gauge.Measurement > 0) ~ Gauge.Elevation + natural_pred  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02,
  
  downwind_propensity_formula = (Gauge.Day.Type == 'Target') ~ Total.Totals + PC1.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure,
  rain_col_name = 'Rain.Gauge.Measurement',
  upwind_subset = Gauge.Day.Type == 'Upwind',
  downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),
  downwind_target_subset = Gauge.Day.Type == 'Target',
  downwind_control_subset = Gauge.Day.Type == 'Control',
  positive_subset = Rain.Gauge.Measurement > 0,
  attr_type = 'ThoEtAl',
  x_downwind_name = c('Gauge.Elevation'),
  
  #Indicator for whether to perform bootstrap inference on attribution and SATE
  bootstrap = TRUE,

  #Specification of various option for bootstrap, e.g., bootstrap_opt(bootstrap_type = 'PREB1'), currently support bootstrap_type = 'PREB0', 'PREB1', 'PREB2', 'REB0', 'REB1', 'REB2', 'MREB1'
  bootstrap_option = bootstrap_opt(B_bootstrap = 500, 
                                   bootstrap_seed = 123, 
                                   bootstrap_parallel = TRUE,
                                   bootstrap_parallel_num_worker = 6),

  #Indicator for whether to perform permutation inference on attribution and SATE
  permutation = TRUE,

  #Specification of various option for permutation,  e.g., whether to permute the operating states between ionizers, between days, between gaugedays
  permutation_option = permutation_opt(B_permutation = 500,
                                       permutation_seed = 999,
                                       permutation_parallel = TRUE,
                                       permutation_parallel_num_worker = 6)
  )
```


The S3 object from `rain_attr` contains detailed results for the bootstrap and permutation analyses, which can be extracted nicely using the `summary` method. In particular, if bootstrap and permutation was performed, the `summary` method would return the bootstrap percentile confidence interval, the bootstrap p-value for positive attribution or SATE (proportion of bootstrapped attribution or SATEs that are less than or equal to zero), and permutation p-value (proportion of permuted attribution or SATEs that are greater than or equal to the estimates obtained from the original data):
```{r, eval=FALSE}
summary(boot_perm_result)
```

The bootstrap and permutation results for the attribution suggest that the increase in the raw-scale rainfall attributable to the ionization technology is positive and significant at the 5% significance level. Similarly, the results for the SATE indicate that the causal effect of ionization technology on downwind rainfall is positive and significant at the 5% level. 

We can plot the bootstrap and permutation distributions of the attribution and SATE using the `plot` method, where solid vertical lines represent estimates based on the original data and dotted vertical lines represent zeros:
```{r, out.width="100%", fig.height= 12, eval=FALSE}
plot(boot_perm_result)
```

## S3 Methods for Objects Returned by rain_attr
The S3 object returned by the main function `rain_attr` is of class `rain_attr`, which supports a wide range of common S3 methods:

* `coef`: computes the fixed effect coefficients and the EBLUPs of random intercepts if the selected model is a LMM, or the regression coefficients if the selected model is a GLM.

* `print`: Print point estimates of attributions and SATEs, along with additional information for the fitted upwind (first-stage) and downwind (second-stage) LMMs.

* `residuals`: Compute residuals of various types for the selected model.

* `fitted` Compute fitted values of the selected model.

* `varcomp`: Compute the random intercept variance and error variance for the different fitted LMMs.

* `predict`: Computed predictions using the selected model.

* `plot`: Plot the bootstrap and permutation distributions of attribution and SATE estimates. Alternatively, can also be used to obtain model diagnostic plots for the selected model.

* `summary`: Returns a list of summary statistics of the two-stage LMM approach, which is an object of class `summary.rain_attr`. This object can be called to print the point estimates of attributions and SATE, along with their bootstrap confidence intervals, bootstrap p-values and permutation p-values if bootstrap and permutation has been performed.

We provide some example usage of these S3 methods below, noting that most of these S3 methods require the selection of a model through the argument `model`, which must be one of "`upwind_lmm`", "`downwind_lmm`" (default), "`downwind_target_lmm`", "`downwind_control_lmm`", "`downwind_logistic`", or "`downwind_propensity`"

Coefficients of the downwind (second-stage) LMM:
```{r, eval=FALSE}
coef(boot_perm_result, model = 'downwind_lmm')$fixef_coef
head(coef(boot_perm_result, model = 'downwind_lmm')$ranef_coef)
```

Residuals of the upwind (first-stage) LMM:
```{r, eval=FALSE}
head(residuals(boot_perm_result, model = 'upwind_lmm', residual_type = 'response', residual_scaled = T))
```

Prediction using the downwind (second-stage) LMM, which consists of both the estimated fixed effects and the EBLUP of the random intercept:
```{r, eval=FALSE}
#Predicting for these two observations:
oman[1:2,]

predict(boot_perm_result, newdata = oman[1:2,], model = 'downwind_lmm', re_include = TRUE, fixef_include = TRUE)
```

Diagnostic plots for downwind (second-stage) LMM:
```{r, out.width="100%", fig.height= 3, eval=FALSE}
plot(boot_perm_result, plot_type = 'model', model = 'downwind_lmm')

```

For more details on these S3 methods applicable to the `rain_attr` class object as well as the returned value from the `summary` method, we refer to `help("rain_attr-class")`.

## Exploratory Data Analysis Plotting Function
The package also provide another function `eda` that performs various exploratory data analyses (EDA) on gauge-day level rainfall enhancement trial data, with the argument `eda_type` controlling the type of EDA:

* `eda_type = num_obs_days`: Compute contigency tables of the number of observations and number of unique days, with rows corresponding to different types of gauge-day observations, and columns corresponding to positive vs. zero rainfall events.

* `eda_type = num_obs_days_by_year`: Same as `eda_type = num_obs_days`, but computed separately for each year.

* `eda_type = hist_day_group_sizes`: Plot histograms of groups sizes for days with positive rainfall using different types of gauge-day observations.

* `eda_type = qq_rain`: Produce Normal Q-Q plots for positive rainfall values (raw or log-transformed) of different types of gauge-day observations.

* `eda_type = ts_by_type`: Plots daily average positive rainfall (raw or log-transformed) by types of gauge-day observations, facetted by year.

* `eda_type = ts_by_gauge`: Plots daily rainfall (raw or log-transformed) time series for each gauge, optionally highlighting a subset of gauges, with faceting by year.

* `eda_type = ts_by_gauge_interactive`: Interactive version of `eda_type = ts_by_gauge` using `plotly`, with points showing gauge identifiers on hover.

* `eda_type = map_static`: Produce a spatial map of annual average positive rainfall (raw or log-transformed). Optional features include overlaying a polygon layer, adding elevation contour lines, and displaying ionizer locations.

* `eda_type = map_dynamic`: Produce an animated map of daily rainfall (raw or log-transformed). The animation can optionally be restricted to a specific year and may include a polygon overlay, elevation contour lines, and ionizer locations.

We provide some example usage of this `eda` function below.


Contingency tables of rainfall events against type of gauge-day observations:
```{r}
contigency_table = eda(eda_type = "num_obs_days",
              data = oman,
              day_column_name = 'TrialDay',
              upwind_subset = Gauge.Day.Type == 'Upwind',
              downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),
              downwind_target_subset = Gauge.Day.Type == 'Target',
              downwind_control_subset = Gauge.Day.Type == 'Control',
              positive_subset = Rain.Gauge.Measurement > 0
)
contigency_table
```

Histogram of group sizes for each trial day with positive rainfall, i.e., the group size of one trial day is defined as the number of gauge-day observations within that day that recorded positive rainfall and satisfied the specific type of gauge-day observation (such as upwind or downwind):
```{r, out.width="100%"}
hist_example = eda(eda_type = "hist_day_group_sizes",
              data = oman,
              day_column_name = 'TrialDay',
              upwind_subset = Gauge.Day.Type == 'Upwind',
              downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),
              positive_subset = Rain.Gauge.Measurement > 0
)
hist_example
```


Daily rainfall (on log-scale) time series for each gauge, with the gauge at the highest elevation highlighted:
```{r, out.width="100%"}
daily_lograin_by_gauge = eda(eda_type = "ts_by_gauge",
              data = oman,
              rain_col_name = 'Rain.Gauge.Measurement',
              day_column_name = 'TrialDay',
              year_column_name = 'Year',
              use_raw = FALSE,
              gauge_id_column_name = 'Gauge.ID',
              ts_focus_gauge = unique(oman$Gauge.ID[oman$Gauge.Elevation == min(oman$Gauge.Elevation)]))
daily_lograin_by_gauge
```

Spatial maps of annual average rainfall (on log scale) facetted by year, with map polygon layer for Oman added using `input_sf = rnaturalearth::ne_countries(scale = "large", country = "Oman", returnclass = "sf")`, ionizer locations plotted using `ionizer_location_df = ionizer_location`, and elevation contour lines added using `elev_contour = TRUE`:
```{r, out.width="100%", fig.height= 6}
annual_map = eda(eda_type = "map_static",
              data = oman,
              rain_col_name = 'Rain.Gauge.Measurement',
              day_column_name = 'TrialDay',
              year_column_name = 'Year',
              use_raw = F,
              longlat_column_names = c("Gauge.Longitude", "Gauge.Latitude"),
              long_lim = c(55,60),
              lat_lim = c(22,25),
              input_sf = rnaturalearth::ne_countries(scale = "large", country = "Oman", returnclass = "sf"),
              ionizer_location_df = ionizer_location,
              ionizer_id_column_name = 'Ionizer',
              ionizer_longlat_column_names = c("Longitude","Latitude"),
              elev_contour = TRUE,
              elev_resolution = 2,
              positive_subset = Rain.Gauge.Measurement > 0
)
annual_map
```


```{r, out.width="100%"}
animated_map = eda(eda_type = "map_dynamic",
              data = oman,
              rain_col_name = 'Rain.Gauge.Measurement',
              day_column_name = 'TrialDay',
              year_column_name = 'Year',
              use_raw = F,
              longlat_column_names = c("Gauge.Longitude", "Gauge.Latitude"),
              long_lim = c(55,60),
              lat_lim = c(22,25),
              input_sf = rnaturalearth::ne_countries(scale = "large", country = "Oman", returnclass = "sf"),
              ionizer_location_df = ionizer_location,
              ionizer_id_column_name = 'Ionizer',
              ionizer_longlat_column_names = c("Longitude","Latitude"),
              elev_contour = TRUE,
              elev_resolution = 2,
              focus_year = c(2013),
              fps = 30
)

animated_map$gif_image
```


For more details on the `eda` function, we refer to `help(eda)`.

## Replicating the Analysis of Chambers et al. (2022a)

## Replicating the Analysis of Chambers et al. (2022b)


