---
title: "Introduction to RainAttr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RainAttr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#library(RainAttr)
devtools::load_all(".")
set.seed(123)
```



## Using the package
This package estimates and performs inference on the **attribution**, defined as the percentage increase in raw-scale rainfall attributable to the ionization technology, and **sample average treatment effect (SATE)** for gauge-day level rainfall enhancement trial data. An example of a gauge-day level dataset is given below:

```{r}
head(oman)
```

The `Gauge.Day.Type` column classifies each observation into one of the following types:

* Upwind: gauges located upwind of ionizers
* Target: gauges located downwind of at least one active ionizers
* Control: gauges located downwind of inactive ionizers
* Out of Scope: gauges that are neither upwind, target, nor control


### Example 

#### Using the main function  

We examine the effectiveness of the ionization technology used in the Oman 2013-2018 trial, based on the two-stage linear mixed model (LMM) approach employed in Chambers et al. (2022a). This is achieved by using the main function `rain_attr` from the package:

```{r, width.cutoff=60}
result = rain_attr(
  #gauge-day level trial data
  data = oman,

  #LMM formula fitted to the upwind (first-stage) observations
  upwind_lmm_formula = LogRain ~  Gauge.Elevation + Steering.Wind.Speed + Total.Totals + PC2.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure + (1|TrialDay),

  #Variable name for storing instrumental prediction from upwind (first-stage) LMM
  instr_pred_name = 'natural_pred',

  #Types of instrumental prediction: 'Unconditional' or 'Conditional'
  instr_pred_type = 'Conditional',

  #LMM formula fitted to the downwind (second-stage) observations
  downwind_lmm_formula = LogRain - natural_pred ~ Gauge.Elevation  + Target.H.01 + Target.H.02 + Target.H.03 + Target.H.04 + Target.H.05 + Target.H.06 + Target.H.07 + Target.H.08 + Target.H.09 + Target.H.10 + Gauge.Elevation:Target.H.01 + Gauge.Elevation:Target.H.02 + (1|TrialDay),

  #Formula for fitting a logistic regression model to the indicator of rainfall event (used for bootstrapping of rainfall events)
  downwind_logistic_formula = NULL,

  #Formula for fitting the propensity score model to downwind observations, where responses are indicators of whether each downwind observations is a 'Target'
  downwind_propensity_formula = (Gauge.Day.Type == 'Target') ~ Total.Totals + PC1.Dry.Temperature + PC1.Relative.Humidity + PC1.Ground.Level.Pressure,

  #Specify the column in input data that contains the raw rainfall
  rain_col_name = 'Rain.Gauge.Measurement',

  #Logical expression identifying subset of observations to which upwind_lmm_formula is fitted
  upwind_subset = Gauge.Day.Type == 'Upwind',

  #Logical expression identifying subset of observations to which downwind_lmm_formula is fitted
  downwind_subset = Gauge.Day.Type  %in% c('Target','Control'),

  #Logical expression identifying subset of observations of those that are 'Target'
  downwind_target_subset = Gauge.Day.Type == 'Target',

  #Logical expression identifying subset of observations of those that are 'Control'
  downwind_control_subset = Gauge.Day.Type == 'Control',

  #Logical expression identifying subset of observations with rainfall event
  positive_subset = Rain.Gauge.Measurement > 0,
  
  #Types of correction (for back-transformation bias) used for computing the attribution estimate: 'ChambersEtAl', 'ChambersEtAl_No_Winsorize', 'ThoEtAl', or 'No'
  attr_type = 'ThoEtAl',

  #Vector of variable names in downwind_lmm_formula to identify non-ionizer related covariates, whose effects are not included in the calculation of attribution and SATE
  x_downwind_name = c('Gauge.Elevation')
  )
```

The `rain_attr` function fits two LMMs sequentially:

1. An upwind (first-stage) LMM fitted using `lme4::lmer(upwind_lmm_formula)` to `data[upwind_subset & positive_subset,]` .
    a) Obtain fitted values (named as `instr_pred_name`) for `data[downwind_subset & positive_subset,]`, where the fitted values are conditional or unconditional depending on `instr_pred_type'.
2. A downwind (second-stage) LMM fitted using `lme4::lmer(downwind_lmm_formula)` to `data[downwind_subset & positive_subset,]`.

The estimated LMM model parameters are then used to calculate attribution and SATE estimates, where `attr_type` specifies the type of correction for back-transformation bias in computing the attribution estimates, and `x_downwind_name` identifies non-ionizer related covariates, whose effects are excluded from the calculation of these estimates.

Moreover, a logistic propensity score model is also fitted using `glm(downwind_propensity_formula, family = "binomial")` to `data[downwind_subset & positive_subset,]` with the response being the binary indicator for whether each observation is exposed to the ionizer (i.e., a target observation). This fitted model, along with the arguments specified via `downwind_target_subset` and `downwind_control_subset`, facilitate the estimation of SATEs.



#### Returned object from the main function

The `rain_attr` function returns an object of class `rain_attr`, which can be called to print the main results associated with the attribution and SATEs. Some basic information related to the estimated upwind (first-stage) and downwind (second-stage) models are also printed, such as the number of observations and the estimated fixed effects.
```{r}
result
```

The `apo` and `apl` attribution estimates indicate that `r round(result$hatattr$apo * 100,2)`% and `r round(result$hatattr$apl * 100,2)`% increase in raw-scale downwind rainfall are attributable to the operation of the ionizers, expressed as a percentage of the total observed rainfall and the total natural rainfall, respectively. We note that attribution estimates are only meaningful when log-rainfall is modeled in both stages, which is typically the case since raw-scale rainfall is often heavily right-skewed, whereas log-rainfall better satisfies the normality assumptions of LMMs.

The above results also show that all SATE estimates are positive, suggesting a causal effect of ionization technology in increasing downwind rainfall. For exact formulas of the five types of SATE estimates, see `help(rain_attr)`. A detailed comparison is also provided in Chambers et al. (2022b).




### Bootstrap and Permutation Inference

## S3 Methods for Objects Returned by rain_attr

## Replicating the Analysis of Chambers et al. (2022a)

## Replicating the Analysis of Chambers et al. (2022b)

## Additional Exploratory Data Analysis Plotting Functions
